Bioverse: The Habitable Zone Inner Edge Discontinuity as an Imprint of Runaway Greenhouse Climates on Exoplanet Demographics

### 2023-07-13 src/tex/ms.tex
address minor comments from referee #1

Yet, \citet{Hamano2013} suggested that the present-day dry conditions on Venus may have been directly inherited from the early magma ocean stage.% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
%This scenario requires early water delivery, which for Venus is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
%Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
For Venus, isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B} and the planet's atmospheric composition suggest that it went through this phase~\citep{2020NatGe..13..265G}.
Although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}, a transient habitable phase cannot be conclusively ruled out~\citep[e.g.,][]{Way2016,Salvador2017,Krissansen-Totton2021}.

@@

\citet{Hardegree-Ullman2023} used the \gaia\ Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
The GCNS identified stars out to 120~pc and includes \gaia\ DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2022} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
From this information, we computed colors and absolute magnitudes and applied initial color-magnitude cuts to remove non-main sequence stars.
We derived effective temperatures primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
Luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
We computed stellar radii with the effective temperatures and luminosities using the Stefan-Boltzmann law or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M~dwarfs.
Finally, we derived masses from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
The derived stellar parameters from \citet{Hardegree-Ullman2023} were compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively.
From this catalog, \bioverse\ samples stars within an isotropic distance from the solar system as required by the planetary sample size.

@@

To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation through water photodissociation and H~escape (see bottem panel of Figure~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
It represents the fraction of planets above the instellation threshold whose atmospheres are currently inflated due to a runaway greenhouse climate.
Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the non-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere.
In the following, we test if and under what conditions this parametrization causes a demographic trend that is large enough to be detected with high significance.

@@

M~dwarf systems are also key for detecting the runaway greenhouse transition:
Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
In fact, we showed that the FGK part of the planet sample barely contributes to the statistical power.
Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars,][]{Turbet2019}, enhancing the demographic signal it leaves.
An additional advantage of targeting M~dwarfs are their extended runaway greenhouse phases, whose duration can reach the order of gigayears~\citep{Luger2015}.
This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
Therefore, in addition to the high scientific value of boosted detections of potentially habitable planets, M dwarfs are also indispensable for the discovery and characterization of the runaway greenhouse transition.
As with a pure volume-limited sample, a targeted M~dwarf survey, too, profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.

>>>>

Yet, \citet{Hamano2013} suggested that the present-day dry conditions on Venus may have been directly inherited from the early magma ocean stage.% if a strongly infrared-absorbing, condensable species such as water is present on the planet.
If a strongly infrared-absorbing, condensable species such as water was dominant in the atmosphere, the resulting strong thermal blanketing effect would prevent the planet to efficiently radiate to space and maintain the surface molten~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
This runaway greenhouse state can extend the magma ocean stage to hundreds of \SI{}{\mega\year}~\citep{2016ApJ...829...63S,2021AsBio..21.1325B}, enough to remove the entire water reservoir from a rocky planet by H$_2$O photolysis and subsequent hydrodynamic escape of hydrogen~\citep{2013ApJ...778..154W,2014ApJ...785L..20W,Luger2015}.
%This scenario requires early water delivery, which for Venus is supported by isotopic and petrological evidence from meteoritic samples~\citep{2018SSRv..214...36A,2022Natur.611..245B}.
%Where the incident stellar flux exceeds a threshold value, a non-linear transition from a mixture of condensed and volatile states to a pure volatile state occurs, and the resulting massive steam atmosphere limits the outgoing radiation from the (molten) planetary surface~\citep{Ingersoll1969,Kasting1988,2010ppc..book.....P,Goldblatt2013,2015ExA....40..449L,2017JGRE..122.1458S}.
For Venus, \rev{the atmospheric composition~\citep{2020NatGe..13..265G} and comprehensive analysis of meteoritic samples across the Solar System~\citep{2018SSRv..214...36A,2022Natur.611..245B}} suggest that it went through this phase.
Although the past presence or absence of a Venusian water ocean has not been definitely established~\citep{Raymond2006,Raymond2007,Hamano2013,Way2016,2019JGRE..124.2015K,2021JGRE..12606643K,Turbet2021,2023PNAS..12009751W}, a transient habitable phase cannot be conclusively ruled out~\citep[e.g.,][]{Way2016,Salvador2017,Krissansen-Totton2021}.

@@

\citet{Hardegree-Ullman2023} used the \gaia\ Catalogue of Nearby Stars~\citep[hereafter GCNS,][]{Smart2021} as the basis for deriving stellar parameters for the \bioverse\ catalog.
The GCNS identified stars out to 120~pc and includes \gaia\ DR3 parallaxes and photometry in $G$, $G_{BP}$, and $G_{RP}$ bands~\citep{GaiaCollaboration2022} and $K_S$ band photometry from 2MASS~\citep{Cutri2003}, all of which were used in stellar classification.
From this information, we computed colors and absolute magnitudes and applied initial color-magnitude cuts to remove non-main sequence stars.
We derived effective temperatures primarily from color-temperature relations derived using the main sequence stellar parameters table from \citet{Pecaut2013}.
Luminosities were computed from absolute $G$-band magnitudes and a derived bolometric correction.
We computed stellar radii with the effective temperatures and luminosities using the Stefan-Boltzmann law or using absolute $K_S$-band magnitudes and an empirical radius-luminosity relation from \citet{Mann2015} for targets within the absolute magnitude range of M~dwarfs.
Finally, we derived masses from the mass-luminosity relation of \citet{Torres2010} for stars with $M_{\star}\gtrsim 0.7\,M_{\odot}$, and from that of \citet{Mann2019} for targets within the absolute magnitude range of M dwarfs.
The derived stellar \rev{parameters were} compared to measured stellar parameters for all known exoplanet hosts from the literature and were found to be consistent within 1\%, 3\%, and 5.5\% for $T_{\mathrm{eff}}$, $R_{\star}$, and $M_{\star}$, respectively, which are all below the typical measurement uncertainties of 3.3\%, 6.8\%, and 7.9\%, respectively~\rev{\citep{Hardegree-Ullman2023}}.
From this catalog, \bioverse\ samples stars within an isotropic distance from the solar system as required by the planetary sample size.

@@

To account for planets unable to sustain a steam atmosphere over extended time spans, as well as evolutionary effects such as desiccation through water photodissociation and H~escape (see \rev{bottom} panel of Figure~\ref{fig:radiusevolution}), we introduce a dilution parameter $f_\mathrm{rgh}$.
It represents the fraction of planets above the instellation threshold whose atmospheres are currently inflated due to a runaway greenhouse climate.
Our simulation setup is such that all planets receiving a net instellation $S < S_\mathrm{thresh}$ follow the non-runaway greenhouse relation, and a fraction $f_\mathrm{rgh}$ of the planets with $S > S_\mathrm{thresh}$ follow the wet melt relation.
The choice $f_\mathrm{rgh} < 1$ comprises all reasons why a planet has no runaway greenhouse climate despite high instellation, for instance the absence of an atmosphere or of volatiles that could form a steam atmosphere.
In the following, we test if and under what conditions this parametrization causes a demographic trend that is large enough to be detected with high significance.

@@

M~dwarf systems are also key for detecting the runaway greenhouse transition:
Our calculations with different spectral types (Sect.~\ref{sec:results_FGK_M}) show that the information content of M~dwarfs in a sample dominates the hypothesis tests.
Besides their large number in a volume/magnitude-limited sample, transiting M~dwarf planets are more likely to be located near the threshold instellation and in particular on orbits further out, i.e., in the optimistic habitable zone.
In fact, we showed that the FGK part of the planet sample barely contributes to the statistical power.
Furthermore, the transit depth difference at the transition is expected to be larger for M~dwarfs~\citep[$\sim \SI{100}{\ppm}$ for early, $\sim \SI{1000}{\ppm}$ for late M~stars,][]{Turbet2019}, enhancing the demographic signal it leaves.
An additional advantage of targeting M~dwarfs are \rev{the extended runaway greenhouse phases of their planets that can last on} the order of gigayears~\citep{Luger2015}.
This increases the probability of observing any given planet in the sample during the runaway greenhouse phase, essentially driving $f_\mathrm{rgh}$ to higher values.
Therefore, in addition to the high scientific value of boosted detections of potentially habitable planets, M dwarfs are also indispensable for the discovery and characterization of the runaway greenhouse transition.
As with a pure volume-limited sample, a targeted M~dwarf survey, too, profits from follow-up measurements of planetary masses with an order of magnitude increase in evidence.
